---
title: "Getting Started with Routing"
---

Packages

```{r}
#| label: set-up-pkgs
pkgs = c(
  "tidyverse",
  "sf",
  "sfnetworks",
  "osmextract",
  "tmap",
  "tmap.networks",
  "tidygraph",
  "igraph",
  "crsuggest"
)

# Installing the packages in the pkgs vector
pak::pkg_install(pkgs)

# Loading all the packages,
# this is equivalent to call library() for each one
sapply(pkgs,require,character.only = TRUE)
```

Loading data from main workbook

```{r}
#| label: load-od-data

if (!file.exists("flows_north_time.rds")) {
  message("Precomputed Northern Spain data not found. Download the raw data with the code block above or download from the GitHub release assets.")
  # Download from GitHub release assets:
  # system("gh release download v1 --pattern 'flows_north_time.rds'")
  release_url <- "https://github.com/tdscience/tartu26/releases/download/v1/"
  for(f in c("flows_north_time.rds", "locations_north.rds")) {
      download.file(paste0(release_url, f), f, mode = "wb")
  }
}

flows_north_time <- readRDS("flows_north_time.rds")
locations_north <- readRDS("locations_north.rds")
```

Extracting OSM data

Sampling four of the most relevant zones

```{r}
#| label: sample-zones
max_flow_zones <- flows_north_time |> 
  slice_max(n = 50,order_by = count) |> 
  select(origin,dest,count) |> 
  pivot_longer(names_to = "type", values_to = "zone",cols = -count) |> 
  mutate(zone = as.character(zone)) |> 
  filter(str_detect(zone,"_|external",negate = TRUE)) |> 
  pull(zone) |> 
  unique()


set.seed(1234)
all_zones <- locations_north |> 
  st_as_sf(coords = c("lon","lat"),crs = 4326)
sample_zones <- locations_north |> 
  st_as_sf(coords = c("lon","lat"),crs = 4326) |>
  filter(id %in% max_flow_zones) |> 
  sample_n(4)

# tmap_mode("view")
# qtm(sample_zones)
  
```

```{r}
#| label: process-network
#| eval: false

zones_buffer <- all_zones |> st_union() |> st_buffer(5e3) |> st_convex_hull()

# Extracting the driving network using osmextract
net <- oe_get_network("Spain",
                      mode = "driving",
                      extra_tags = c("maxspeed")
                      # boundary = zones_buffer,
                      # boundary_type = "clip"
                      )

highway_types_valid <-  c("motorway","trunk","primary","secondary")

# Simplifying the highway tag
net$highway <- str_remove(net$highway,"_link")

# net$highway |> unique()

# filtering only relevant road types
net <- net |> filter(highway %in% highway_types_valid)

# Casting the sf object as a compatible geometry for sfnetworks
net <- net |> st_cast("LINESTRING")

# eliminating the other_tags column
net$other_tags <- NULL


# Extracting the most frequent speed limits based on distance of elements 
speed_limits <- net |>
  mutate(dist = as.numeric(st_length(geometry))) |> 
  st_drop_geometry() |>
  drop_na(maxspeed) |>
  summarise(across(dist,sum),.by = c(highway,maxspeed)) |>
  slice_max(dist,by = "highway",n = 1)

# Creating a column for filling the speed limits if missing
net$maxspeed_ifna <- as.numeric(speed_limits$maxspeed[match(net$highway,speed_limits$highway)])

# Casting the maxspeed as numeric
net$maxspeed <- as.numeric(net$maxspeed)

# Filling missing speed limits
net$maxspeed <- coalesce(net$maxspeed,net$maxspeed_ifna)

# Saving 
net <- net |> 
  select(osm_id,
         name,
         highway,
         maxspeed)

zones_mask <- st_within(net,zones_buffer) |> vapply(length,integer(1))
zones_mask <- zones_mask>0
# Saving the clean network for the whole contry
write_rds(net,"clean_net.rds")
write_rds(net[zones_mask,],"clean_net_study_area.rds")

system("gh release upload v1 clean_net.rds clean_net_study_area.rds --clobber")

```



```{r}
#| lable: load-clean-sf-net

if(!file.exists("clean_net_study_area.rds")){
  message("Precomputed Northern Spain network not found. Download the raw data with the code block below or download from the GitHub release assets.")
  
  release_url <- "https://github.com/tdscience/tartu26/releases/download/v1/"
  for(f in c("clean_net_study_area.rds")) {
      download.file(paste0(release_url, f), f, mode = "wb")
  }
}

net <- readRDS("clean_net_study_area.rds")
```

Creating the primal graph (undirected) from the sf object using `sfnetworks`

```{r}
# Identifying the best crs for our data

best_crs <- suggest_top_crs(net, units = "m")

# Creating the sfnetwork
sfnet <- net |>
  st_transform(best_crs) |> 
  as_sfnetwork(directed = FALSE)

# Preprocessing
## Dividing the linestrings using overlapping points creating junctions
sf_net_subdiv <- convert(sfnet, to_spatial_subdivision)


## Calculating travel times for each edge based on speed limit
sf_net_subdiv <- sf_net_subdiv |>
  activate(nodes) |> 
  filter(group_components() == 1) |> 
  activate(edges) |> 
  mutate(speed = units::set_units(maxspeed[cur_group_id()], "km/h"),
         dist = st_length(geometry),
         tra_time = units::set_units(dist/(speed),"min"))

## Removing interstitial nodes (nodes degree 2)
sf_net_simpl <- convert(sf_net_subdiv,
                        to_spatial_smooth,
                        summarise_attributes = list(dist = "sum",
                                                    tra_time = "sum",
                                                    osm_id = "first",
                                                    highway = "first"))
```

A quick visualisation

```{r}
tm_shape(sf_net_simpl)+
  tm_edges("highway")+
  tm_shape(sample_zones)+
  tm_dots("red")
```

Some metrics about our network graph

```{r}
# Diameter in travel time
diameter(sf_net_simpl,directed = FALSE, weights = pull(sf_net_simpl,tra_time))

#Diameter in distance
diameter(sf_net_simpl,directed = FALSE, weights = pull(sf_net_simpl,dist))*1e-3
```

Exploring the centrality of the network

```{r}
sf_net_simpl <- sf_net_simpl |> 
  activate(edges) |> 
  mutate(bc = centrality_edge_betweenness(weights = tra_time))

ggplot() +
  geom_sf(data = st_as_sf(sf_net_simpl, "edges"), aes(col = log10(bc)))+
  scale_color_continuous(palette = "viridis")+
  theme_void()
```

Extracting routes for the selected zones

```{r}
sf_nodes <- st_as_sf(sf_net_simpl,"nodes")
node_indexes <- sample_zones |>
  st_transform(st_crs(sf_net_simpl)) |> 
  st_nearest_feature(sf_nodes)

sample_od <- expand.grid(origin = node_indexes[1],
                         dest = node_indexes) |>
  filter(origin != dest)
```

Extracting paths

```{r}
# code extracted from sfnetworks documentation

paths = st_network_paths(sf_net_simpl, from = node_indexes[1], to = sample_od$dest, weights = "tra_time")
paths

plot_path = function(node_path) {
  sf_net_simpl %>%
    activate("nodes") %>%
    slice(node_path) %>%
    plot(cex = 1.5, lwd = 1.5, add = TRUE)
}

colors = sf.colors(4, categorical = TRUE)

plot(sf_net_simpl, col = "grey")
paths %>%
  pull(node_paths) %>%
  walk(plot_path)
sf_net_simpl %>%
  activate("nodes") %>%
  st_as_sf() %>%
  slice(node_indexes) %>%
  plot(col = colors, pch = 8, cex = 2, lwd = 2, add = TRUE)

```

Extracting distance and travel time for each OD pair

```{r}

```

